### 2021-08-20 manuscript/ex_article.tex
Submitted

The first part of our developments concerns monitoring the species in the system and estimating the deviation from the true population density function. We do this as follows: Since each species is described by an abundance-trait-covariance tuple, we map these to time dependent distributions. Then, using derived a posteriori error estimates of the population level model, we track the modeling error of each species' associated distribution function. The second part concerns the case of residual blow-up, which implies a model breakdown at the species level, here associated with a speciation event. Concerning the actual speciation event we compare two different approaches with regards to splitting the parent species into new child species;
(1) \emph{Heuristic approach}: When a residual blow-up is detected, we proceed to split the relevant reconstructed distribution function along the trait directions orthogonal to the direction of divergence. Each of these sub-distributions are then mapped back to abundance-trait-variance tuples, thus yielding the new child species to be incorporated at the species level. (2) \emph{Multi-scale approach}: When a residual blow-up is detected, we delegate the relevant species' reconstructed density function to the population level model to be solved in an appropriate local region of trait space while coupled to the species level model which governs the interaction of the remaining species in the system. Simultaneously, we map the local abundance density function to  abundance-trait-variance tuples and measure the distance between the mean trait coordinates. At such time when the new child species are sufficiently separated in trait space (e.g., by a multiple of the maximum trait standard deviation), we incorporate the new child species at the species level and decouple the multi-scale model. We assess the accuracy of each approach by comparing the species parameters (pre- and post-speciation) to the corresponding moments of the reference (global) PLM solution.

@@

\begin{figure}[h]
\centering
  \includegraphics[width=.6\linewidth]{figures/jan_fig}
 \caption{Coupling of SLM and PLM during speciation events. Shaded area in blue indicates the macro-scale (SLM) model is solved (here, the curve represents the reconstructed distribution function associated with each species' abundance-trait-covariance tuple), the green area indicates the micro-scale (PLM) model is solved (here, the curve represents the abundance density, or, solution function of the PLM).}
\label{fig:jan}
\end{figure}

@@

Find $N :(0,T) \rightarrow \real^{s(1+d+d^2)}$ such that $n_i \geq 0$, for $1\leq i \leq s$, and 
\bse
\begin{alignat}{2}
\dfrac{\textnormal{d}N}{\textnormal{d}t} &= \Ncal(N), \quad && t\in(0,T), \label{ODEmodel}\\
N &= N_0, &&t = 0. \label{ODEmodelIC}
\end{alignat}
\ese
In practice, the problem \eqref{ODEmodel}--\eqref{ODEmodelIC} can be reduced to $s(1+d+d(d+1)/2)$ equations due to the symmetry of the trait covariance matrix. For additional details including a derivation of the above model, see~\cite{nordbotten2020dynamics}. Note also that when indexing the components of a point $x\in\Omega$ we use subscripts, i.e., $x = (x_1,\cdots,x_d)$, not to be confused with the species mean trait coordinate $x_i$, where we write $x_i(t) = (x_{i,1}(t),\cdots,x_{i,d}(t))$. 
%The distinction should be clear from the context.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Relationship between population and species level models}
If there is a time interval $J := (t_{0},t_1) \subset (0,T)$ where $s \geq 1$ species can unambiguously be identified, there exists a collection of non-overlapping (possibly time dependent), open and bounded regions in trait space, i.e., $\{B_i \subset \Omega, i \leq s,  B_i \cap B_j = \emptyset, i \neq j\}$, where the population density is compactly supported and contains only one local maximum. Thus, if we let $\mu_{B_i}^k(n)$, be the $k$'th order moment of $n$ (for $k \in \{0,1,2\}$, \red{and assuming $n$ is smooth enough for these moments to exist}), within the region $B_i$, i.e.,
\bse
\begin{alignat}{1}
\mu_{B_i}^0(n) &:= \int_{B_i} n(x)\dx,\label{m0}\\ 
\mu_{B_i}^1(n) &:= \frac{1}{\mu_{B_i}^0(n)} \int_{B_i} n(x)x\dx,\label{m1}\\
\mu_{B_i}^2(n) &:= \frac{1}{\mu_{B_i}^0(n)} \int_{B_i} n(x)(x-\mu_{B_i}^1)\otimes (x-\mu_{B_i}^1)\dx,\label{m2}
\end{alignat}
\ese
we can initialize the SLM at $t=t_{0}$ with initial data
\begin{equation}
N_{0} = (\mu_{B_{1}}^0,\mu_{B_{1}}^1,\mu_{B_{1}}^2,\cdots,\mu_{B_{s}}^0,\mu_{B_{s}}^1,\mu_{B_{s}}^2)|_{t=t_0}.
\end{equation}
Solving \eqref{ODEmodel}--\eqref{ODEmodelIC} then amounts to approximating \eqref{m0}--\eqref{m2} for $t \in J$.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Time discrete PLM}\label{sec:timedisc}
In this section we introduce the time discrete PLM, which will serve as the basis for the a posteriori error bounds: 
\red{
For $k = 0,\cdots,M$, we let $t_k$ denote the discrete times, such that $0 = t_0 < t_1 < \cdots < t_M := T$, and where $t_k = t_{k-1} + \tau_k = \sum_{j = 1}^k \tau_k$, for some set of time increments $\{\tau_k\}_{k=1}^M$ (in the following, we use superscripts to indicate dependency on the discrete times, e.g., $u^k := u(t_k)$). The time-discrete version of eq.~\ref{PDE} then reads as (for $k\geq 1$)
\begin{equation}
n^k - r^k n^k + \Phi^k(n^k)n^k - \nabla \cdot (g^k \nabla n^k) = f + n^{k-1},\label{timediscPDE}
\end{equation}
where we have discretized in time using a standard implicit first order method (i.e., the backward Euler method). While we do not prove convergence of the above discrete scheme to the original continuous problem, the local truncation error for the backward Euler method is known to be of second order, and one would expect first-order convergence of the time-discrete problem to the continuous problem for sufficiently smooth problems. }

@@

In this section we present our strategies for the actual speciation event. The first is based on splitting the support of the relevant reconstructed density function into subregions of trait space, and mapping back to new species abundance-trait-variance tuples by calculating the moments of the reconstructed density in each subregion. The second approach is based on coupling the PLM and SLM in a multi-scale framework for the duration of the speciation event, where the diverging species is delegated to the PLM to be solved locally in trait space.

@@

\pt n^{\ell} - rn^\ell + \Phi(n^\ell)n^{\ell} - \nabla \cdot (g \nabla n^\ell) + n^\ell\Cbb(N^{\ell}) = f^\ell, &\textn{ in } B_{i,\nu}^k, \, \ell > k \label{HMM2} \\

@@

\dfrac{\textnormal{d}N^\ell}{\textnormal{d}t} = \Ncal(N^\ell) + \Dbb(n^\ell)N^\ell, &\ell > k \label{HMM1} \\[1ex]

@@

where $\pt n^{\ell}$ and $\dfrac{\textnormal{d}N^\ell}{\textnormal{d}t}$ denotes the discrete time derivatives for the PLM and SLM, respectively, and where $\Cbb : \real^{s(1+d+d^2)} \rightarrow \real$ and $\Dbb : H_{0,+}^1(B_{i,\nu}^k) \rightarrow \real^{s(1+d+d^2)} $ are the micro/macro couplings, approximating respectively the individual and species interactions. These are defined as follows: 

@@

For both examples we assess the accuracy of the speciation methods by solving simultaneously the PLM globally and calculating the moments which are then compared to the corresponding species parameters. Moments of the reference PLM solution are always shown as dotted lines, and indicated by the letter $\mu$ in the legends. For the multi-scale method, the start and end of the speciation event is indicated by two vertical dotted lines. For the heuristic method, speciation is indicated by a single vertical dotted line. 
%Furthermore, for the second numerical example the monotonicity condition \eqref{monotonicity} is not fulfilled, and hence the assertion of Lemma~\ref{lem:edist} is not true. Nonetheless, from a modeling point of view the multi-scale framework developed is still applicable.

>>>>

The first part of our developments concerns monitoring the species in the system and estimating the deviation from the true population density function. We do this as follows: Since each species is described by an abundance-trait-covariance tuple, we map these to time dependent distributions. Then, using derived a posteriori error estimates of the population level model, we track the modeling error of each species' associated distribution function. The second part concerns the case of residual blow-up, which implies a model breakdown at the species level, here associated with a speciation event. Concerning the actual speciation event we compare two different approaches with regards to splitting the parent species into new child species;
(1) \emph{Heuristic approach}: When a residual blow-up is detected, we proceed to split the relevant reconstructed distribution function along the trait directions orthogonal to the direction of divergence. Each of these sub-distributions are then mapped back to abundance-trait-variance tuples, thus yielding the new child species to be incorporated at the species level. (2) \emph{Multi-scale approach}: When a residual blow-up is detected, we delegate the relevant species' reconstructed density function to the population level model to be solved in an appropriate local region of trait space while coupled to the species level model which governs the interaction of the remaining species in the system. Simultaneously, we map the local abundance density function to  abundance-trait-variance tuples and measure the distance between the mean trait coordinates. At such time when the new child species are sufficiently separated in trait space (e.g., by a multiple of the maximum trait standard deviation), we incorporate the new child species at the species level and decouple the multi-scale model. We assess the accuracy of each approach by comparing the species parameters (pre- and post-speciation) to the corresponding \red{statistical} moments of the reference (global) PLM solution.

@@

\begin{figure}[h]
\centering
  \includegraphics[width=.6\linewidth]{figures/jan_fig}
 \caption{Coupling of SLM and PLM during speciation events. \red{Outside the green area} the macro-scale (SLM) model is solved (here, the curve represents the reconstructed distribution function associated with each species' abundance-trait-covariance tuple), the green area indicates the micro-scale (PLM) model is solved (here, the curve represents the abundance density, or, solution function of the PLM).}
\label{fig:jan}
\end{figure}

@@

Find $N :(0,T) \rightarrow \real^{s(1+d+d^2)}$ such that $n_i \geq 0$, for $1\leq i \leq s$, and 
\bse
\begin{alignat}{2}
\dfrac{\textnormal{d}N}{\textnormal{d}t} &= \Ncal(N), \quad && t\in(0,T), \label{ODEmodel}\\
N &= N_0, &&t = 0. \label{ODEmodelIC}
\end{alignat}
\ese
In practice, the problem \eqref{ODEmodel}--\eqref{ODEmodelIC} can be reduced to $s(1+d+d(d+1)/2)$ equations due to the symmetry of the trait covariance matrix. For additional details including a derivation of the above model, see~\cite{nordbotten2020dynamics}. Note also that when indexing the components of a point $x\in\Omega$ we use subscripts, i.e., $x = (x_1,\cdots,x_d)$, not to be confused with the species mean trait coordinate $x_i$, where we write $x_i(t) = (x_{i,1}(t),\cdots,x_{i,d}(t))$. 
%The distinction should be clear from the context.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Relationship between population and species level models}
If there is a time interval $J := (t_{0},t_1) \subset (0,T)$ where $s \geq 1$ species can unambiguously be identified, there exists a collection of non-overlapping (possibly time dependent), open and bounded regions in trait space, i.e., $\{B_i \subset \Omega, i \leq s,  B_i \cap B_j = \emptyset, i \neq j\}$, where the population density is compactly supported and contains only one local maximum. Thus, if we let $\mu_{B_i}^k(n)$, be the $k$'th order moment of $n$ (for $k \in \{0,1,2\}$, \red{and assuming $n$ is smooth enough for these moments to exist}), within the region $B_i$, i.e.,
\bse
\begin{alignat}{1}
\mu_{B_i}^0(n) &:= \int_{B_i} n(x)\dx,\label{m0}\\ 
\mu_{B_i}^1(n) &:= \frac{1}{\mu_{B_i}^0(n)} \int_{B_i} n(x)x\dx,\label{m1}\\
\mu_{B_i}^2(n) &:= \frac{1}{\mu_{B_i}^0(n)} \int_{B_i} n(x)(x-\mu_{B_i}^1)\otimes (x-\mu_{B_i}^1)\dx,\label{m2}
\end{alignat}
\ese
we can initialize the SLM at $t=t_{0}$ with initial data
\begin{equation}
N_{0} = (\mu_{B_{1}}^0,\mu_{B_{1}}^1,\mu_{B_{1}}^2,\cdots,\mu_{B_{s}}^0,\mu_{B_{s}}^1,\mu_{B_{s}}^2)|_{t=t_0}.
\end{equation}
Solving \eqref{ODEmodel}--\eqref{ODEmodelIC} then amounts to approximating \eqref{m0}--\eqref{m2} for $t \in J$.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Time discrete PLM}\label{sec:timedisc}
In this section we introduce the time discrete PLM, which will serve as the basis for the a posteriori error bounds: 
\red{
For $k = 0,\cdots,M$, we let $t_k$ denote the discrete times, such that $0 = t_0 < t_1 < \cdots < t_M := T$, and where $t_k = t_{k-1} + \tau_k = \sum_{j = 1}^k \tau_k$, for some set of time increments $\{\tau_k\}_{k=1}^M$ (in the following, we use superscripts to indicate dependency on the discrete times, e.g., $u^k := u(t_k)$). The time-discrete version of eq.~\ref{PDE} then reads as (for $k\geq 1$)
\begin{equation}
n^k - \tau_k r^k n^k + \tau_k \Phi^k(n^k)n^k - \tau_k \nabla \cdot (g^k \nabla n^k) = \tau_k f + n^{k-1},\label{timediscPDE}
\end{equation}
where we have discretized in time using a standard implicit first order method (i.e., the backward Euler method). While we do not prove convergence of the above discrete scheme to the original continuous problem, the local truncation error for the backward Euler method is known to be of second order, and one would expect first-order convergence of the time-discrete problem to the continuous problem for sufficiently smooth problems. }

@@

In this section we present our strategies for the actual speciation event. The first is based on splitting the support of the relevant reconstructed density function into subregions of trait space, and mapping back to new species abundance-trait-variance tuples by calculating the \red{statistical} moments of the reconstructed density in each subregion. The second approach is based on coupling the PLM and SLM in a multi-scale framework for the duration of the speciation event, where the diverging species is delegated to the PLM to be solved locally in trait space.

@@

\pt n^{\ell} - rn^\ell + \Phi(n^\ell)n^{\ell} - \nabla \cdot (g \nabla n^\ell) + n^\ell\Cbb^\ell(N^{\ell}) = f^\ell, &\textn{ in } B_{i,\nu}^k, \, \ell > k \label{HMM2} \\

@@

\dfrac{\textnormal{d}N^\ell}{\textnormal{d}t} = \Ncal(N^\ell) + \Dbb^\ell(n^\ell)N^\ell, &\ell > k \label{HMM1} \\[1ex]

@@

where $\pt n^{\ell}$ and $\dfrac{\textnormal{d}N^\ell}{\textnormal{d}t}$ denotes the discrete time derivatives for the PLM and SLM, respectively, and where $\Cbb^\ell : \real^{s(1+d+d^2)} \rightarrow \real$ and $\Dbb^\ell : H_{0,+}^1(B_{i,\nu}^k) \rightarrow \real^{s(1+d+d^2)} $ are the micro/macro couplings, approximating respectively the individual and species interactions. These are defined as follows: 

@@

For both examples we assess the accuracy of the speciation methods by solving simultaneously the PLM globally and calculating the \red{statistical} moments which are then compared to the corresponding species parameters. Moments of the reference PLM solution are always shown as dotted lines, and indicated by the letter $\mu$ in the legends. For the multi-scale method, the start and end of the speciation event is indicated by two vertical dotted lines. For the heuristic method, speciation is indicated by a single vertical dotted line. 
%Furthermore, for the second numerical example the monotonicity condition \eqref{monotonicity} is not fulfilled, and hence the assertion of Lemma~\ref{lem:edist} is not true. Nonetheless, from a modeling point of view the multi-scale framework developed is still applicable.
